Cooling Requirements for the Vertical Shear Instability in Protoplanetary Disks

### 2015-04-23 paper.tex
messed up notation change. need to fix by inspection.

\begin{abstract}
  It is difficult to understand how cold circumstellar disks accrete onto their central stars.  Due to 
  lower levels of ionization, the magnetic transport mechanisms that
  operate in warmer disks are less effective.  
  A hydrodynamic mechanism, the vertical shear instability (VSI), offers a means to drive angular momentum 
  transport in cold accretion disks such as protoplanetary disks (PPDs).  
  The VSI is driven by a weak vertical gradient in the disk's orbital motion.
  In order to grow, the VSI must overcome vertical buoyancy, a strong
  stabilizing influence in cold disks, where  
  heating is dominated by external irradiation.  Rapid cooling, via
  radiative losses, reduces the effective buoyancy 
  and allows the VSI to operate.  In this paper, we quantify the
  cooling timescale, $t_c$, needed for growth of the VSI 
  and consider the implications for PPDs where $t_c$ is determined by the opacity of dust 
  grains.  We perform a linear analysis of the VSI with cooling in
  vertically global and radially local disk models. 
  For irradiated disks, we find that the VSI is most vigorous for
  rapid cooling with $t_c < \OmK^{-1} \epsilon |q| / (\gamma -1)$ 
  in terms of the Keplerian orbital frequency, $\OmK$; the disk's
  aspect ratio, $\epsilon \ll 1$; the radial  
  powerlaw temperature gradient, $q$; and the adiabatic index,
  $\gamma$.  For longer cooling times, the VSI  
  is much less effective because growth slows and shifts to smaller length scales, which are more prone to 
  viscous or turbulent decay.  We apply our results to a  Minimum Mass Solar Nebula disk model.  We
  find that the VSI is most effective at intermediate radii, from $\sim 5$AU to $\sim 50$AU with a
  characteristic growth time of $\sim 30$ local orbital periods.
  Growth is suppressed by long cooling times both in the opaque  
  inner disk and the optically thin outer disk.  A reduction in the
  dust opacity by a factor of 10 increases cooling times  
  enough to quench the VSI at all disk radii.  Thus the formation of
  solid protoplanets, a sink for dust grains, can impede the VSI. 
\end{abstract}

@@

\section{Thermodynamic requirement for the vertical shear instability}\label{vsi_require}     In this paper we will find that in a stably stratified thin disk, the VSI  
requires a thermal timescale $t_c$ much less than the
disk dynamical timescale, 
\begin{align}
  t _c \ll \Omega_k^{-1}. 
\end{align}
where $\Omega_k$ is the Keplerian frequency. 
To see why this should be expected, we need to compare vertical shear
with vertical buoyancy. For this preliminary discussion we consider a
vertically isothermal disk with aspect-ratio $\epsilon$,  
dimensionless radial temperature gradient $q$ and adiabatic index
$\gamma$. For a PPD, $\epsilon \sim 0.05$, $|q|\sim 1$ and $\gamma\simeq 1.4$.

For a thin disk, the specific angular momentum varies with height as 
\begin{align}\label{vshear_thin}
  r \frac{\p \Omega^2 }{\p z} \simeq \epsilon q \Omega_k^2\left(\frac{z}{H}\right),
\end{align}
where $H$ is the disk scale height. As noted in
\S\ref{intro} this vertical variation is a source of free energy and
thus vertical shear ($\p_z\Omega \neq0$) is destabilizing.

For $\gamma>1$ the disk is stably stratified with a real vertical
buoyancy frequency $N_z$.  In a thin disk,     
\begin{align}\label{nz_thin}
  N_z^2 \simeq \left(\frac{\gamma-1}{\gamma}\right) \left(\frac{z}{H}\right)^2
  \Omega_k^2.  
\end{align}
This provides a stabilizing effect as buoyancy opposes vertical
motion, which is required for the VSI. In fact, if we consider
axisymmetric adiabatic disturbances, then because the 
Solberg-Hoiland criteria \citep{tassoul78} are usually satisfied in a
thin disk, the disk would be stable.

@@

How short must $t_c$ be for vertical shear to prevail? A rough 
comparison of vertical shear to buoyancy for non-adiabatic
perturbations can be obtained by applying Eq. 47 of \cite{urpin03}, which is based on 
\cite{townsend58},   
\begin{align}\label{bcrit_est}
  N_z^2 t_c \lesssim \left|r\frac{\p\Omega}{\p z}\right|. 
\end{align}
The left-hand side represents an effective buoyancy, which is
stabilizing; while the right-hand side represents vertical shear,
which is destabilizing. In a thin disk $\epsilon \ll 1$, so the destabilizing effect of 
vertical shear is rather weak because $|r\p_z\Omega|\sim\epsilon\Omega_k
\ll \Omega_k$. By comparison, the buoyancy frequency $N_z\sim
\Omega_k$. It follows that $t_c\lesssim O(\epsilon\Omega_k^{-1})$ is
required for instability.  \emph{In a thin  
  disk the thermal timescale must be shorter than the dynamical
  timescale by $O(\epsilon)\ll 1$ for the VSI to operate.}

@@

Our central result is the requirement 
\begin{align}\label{prelim_bcrit}
  (\gamma - 1)t_c < \epsilon |q| \Omega_k^{-1}
\end{align}   
for the VSI in a vertically isothermal, thin disk. This equation
reflects Eq. \ref{bcrit_est} but there is no $z$ dependence. 
% While assumptions and simplifications will be made 
% to derive this, 
Eq. \ref{prelim_bcrit} can be 
readily applied to evaluate where in a disk the VSI is likely to
operate. For example, we show in Fig. \ref{bcrit_mmsn2d} an estimate
of the thermal timescales the PPD model shown in
Fig. \ref{eqm_structure}. Here $k_x$ is a radial
wavenumber of the disturbance being considered, thus larger $k_x$
correspond to smaller perturbations. The region outside the red line
satisfies Eq. \ref{prelim_bcrit}.

>>>>

\begin{abstract}
  It is difficult to understand how cold circumstellar disks accrete onto their central stars.  Due to 
  lower levels of ionization, the magnetic transport mechanisms that
  operate in warmer disks are less effective.  
  A hydrodynamic mechanism, the vertical shear instability (VSI), offers a means to drive angular momentum 
  transport in cold accretion disks such as protoplanetary disks (PPDs).  
  The VSI is driven by a weak vertical gradient in the disk's orbital motion.
  In order to grow, the VSI must overcome vertical buoyancy, a strong
  stabilizing influence in cold disks, where  
  heating is dominated by external irradiation.  Rapid cooling, via
  radiative losses, reduces the effective buoyancy 
  and allows the VSI to operate.  In this paper, we quantify the
  cooling timescale, $t_c$, needed for growth of the VSI 
  and consider the implications for PPDs where $t_c$ is determined by the opacity of dust 
  grains.  We perform a linear analysis of the VSI with cooling in
  vertically global and radially local disk models. 
  For irradiated disks, we find that the VSI is most vigorous for
  rapid cooling with $t_c < \OmK^{-1}  h |q| / (\gamma -1)$ 
  in terms of the Keplerian orbital frequency, $\OmK$; the disk's
  aspect ratio, $ h \ll 1$; the radial  
  powerlaw temperature gradient, $q$; and the adiabatic index,
  $\gamma$.  For longer cooling times, the VSI  
  is much less effective because growth slows and shifts to smaller length scales, which are more prone to 
  viscous or turbulent decay.  We apply our results to a  Minimum Mass Solar Nebula disk model.  We
  find that the VSI is most effective at intermediate radii, from $\sim 5$AU to $\sim 50$AU with a
  characteristic growth time of $\sim 30$ local orbital periods.
  Growth is suppressed by long cooling times both in the opaque  
  inner disk and the optically thin outer disk.  A reduction in the
  dust opacity by a factor of 10 increases cooling times  
  enough to quench the VSI at all disk radii.  Thus the formation of
  solid protoplanets, a sink for dust grains, can impede the VSI. 
\end{abstract}

@@

\section{Thermodynamic requirement for the vertical shear instability}\label{vsi_require}     In this paper we will find that in a stably stratified thin disk, the VSI  
requires a thermal timescale $t_c$ much less than the
disk dynamical timescale, 
\begin{align}
  t _c \ll \Omega_K^{-1}. 
\end{align}
where $\Omega_K$ is the Keplerian frequency. 
To see why this should be expected, we need to compare vertical shear
with vertical buoyancy. For this preliminary discussion we consider a
vertically isothermal disk with aspect-ratio $ h$,  
dimensionless radial temperature gradient $q$ and adiabatic index
$\gamma$. For a PPD, $ h \sim 0.05$, $|q|\sim 1$ and $\gamma\simeq 1.4$.

For a thin disk, the specific angular momentum varies with height as 
\begin{align}\label{vshear_thin}
  r \frac{\p \Omega^2 }{\p z} \simeq  h q \Omega_K^2\left(\frac{z}{H}\right),
\end{align}
where $H$ is the disk scale height. As noted in
\S\ref{intro} this vertical variation is a source of free energy and
thus vertical shear ($\p_z\Omega \neq0$) is destabilizing.

For $\gamma>1$ the disk is stably stratified with a real vertical
buoyancy frequency $N_z$.  In a thin disk,     
\begin{align}\label{nz_thin}
  N_z^2 \simeq \left(\frac{\gamma-1}{\gamma}\right) \left(\frac{z}{H}\right)^2
  \Omega_K^2.  
\end{align}
This provides a stabilizing effect as buoyancy opposes vertical
motion, which is required for the VSI. In fact, if we consider
axisymmetric adiabatic disturbances, then because the 
Solberg-Hoiland criteria \citep{tassoul78} are usually satisfied in a
thin disk, the disk would be stable.

@@

How short must $t_c$ be for vertical shear to prevail? A rough 
comparison of vertical shear to buoyancy for non-adiabatic
perturbations can be obtained by applying Eq. 47 of \cite{urpin03}, which is based on 
\cite{townsend58},   
\begin{align}\label{bcrit_est}
  N_z^2 t_c \lesssim \left|r\frac{\p\Omega}{\p z}\right|. 
\end{align}
The left-hand side represents an effective buoyancy, which is
stabilizing; while the right-hand side represents vertical shear,
which is destabilizing. In a thin disk $ h \ll 1$, so the destabilizing effect of 
vertical shear is rather weak because $|r\p_z\Omega|\sim h\Omega_K
\ll \Omega_K$. By comparison, the buoyancy frequency $N_z\sim
\Omega_K$. It follows that $t_c\lesssim O( h\Omega_K^{-1})$ is
required for instability.  \emph{In a thin  
  disk the thermal timescale must be shorter than the dynamical
  timescale by $O( h)\ll 1$ for the VSI to operate.}

@@

Our central result is the requirement 
\begin{align}\label{prelim_bcrit}
  (\gamma - 1)t_c <  h |q| \Omega_K^{-1}
\end{align}   
for the VSI in a vertically isothermal, thin disk. This equation
reflects Eq. \ref{bcrit_est} but there is no $z$ dependence. 
% While assumptions and simplifications will be made 
% to derive this, 
Eq. \ref{prelim_bcrit} can be 
readily applied to evaluate where in a disk the VSI is likely to
operate. For example, we show in Fig. \ref{bcrit_mmsn2d} an estimate
of the thermal timescales the PPD model shown in
Fig. \ref{eqm_structure}. Here $k_x$ is a radial
wavenumber of the disturbance being considered, thus larger $k_x$
correspond to smaller perturbations. The region outside the red line
satisfies Eq. \ref{prelim_bcrit}.

### 2015-04-23 setup.tex
messed up notation change. need to fix by inspection.

\section{Linear problem}\label{linear}
We consider axisymmetric perturbations to the above equilibria in the
form 
\begin{align}
  \rho \to \rho(r, z) + \delta\rho(r,z)\exp{\left( - \ii\sigma
      t\right)},    
\end{align}
and similarly for other fluid variables, where $\sigma = \omega + \ii
\nu$ is a (generally) complex frequency, with $\omega$ and $\nu$ being
the real frequency and growth rate, respectively.

The linearized fluid equations in cylindrical co-ordinates are  
\begin{align}
  \ii\sigma\delta\rho &= \frac{\rho}{r}\frac{\p}{\p r}\left(r\delta v_r\right) + \frac{\p}{\p
    z}\left(\rho\delta v_z\right)  + \hat{g}_c\frac{\p\rho}{\p r}\delta v_r,\label{full_mass1}\\
  \ii\sigma\delta v_r &= \frac{1}{\rho}\frac{\p\delta P}{\p r} - 2\Omega
  \delta v_\phi - \hat{g}_c\frac{\delta\rho}{\rho^2}\frac{\p P}{\p r},\\ 
  \ii\sigma\delta v_\phi  &= \frac{\kappa^2}{2\Omega}\delta v_r +
  r\frac{\p\Omega}{\p z}\delta v_z,\\
  \ii \sigma\delta v_z &= \frac{1}{\rho}\frac{\p\delta P}{\p z} -
  \frac{\delta \rho}{\rho^2}\frac{\p P}{\p z},\\
  \ii\sigma\delta P &=\frac{\p P}{\p z}\delta v_z + \gamma P
  \left[\frac{1}{r}\frac{\p}{\p r}\left(r\delta v_r\right) + \frac{\p
      \delta v_z}{\p z}\right]\notag \\
  &+ \frac{1}{t_c}\left(\delta P -  
    \frac{P}{\rho}\delta\rho\right) + \hat{g}_c \frac{\p P}{\p r}\delta v_r,\label{full_energy1}
\end{align}
where the coefficients $\hat{g}_c=1$ are inserted to label terms
associated with the background radial disk structure.

@@

\subsection{Radially-local approximation}  
We consider perturbations with small radial length-scales 
so that we can ignore the radial variations in the background
structure and radial gradients. \citetalias{barker15} have validated this
approximation by comparison with the linear VSI in global cylindrical
disk geometry. We thus evaluate all coefficients in
Eq. \ref{full_mass1}---\ref{full_energy1} at the fiducial radius
$r=r_0$. However, the full vertical dependence of the coefficients are
retained. Then we can assume perturbations have spatial dependence in
the form 
\begin{align}
  \delta \rho(r,z) \to \delta\rho(z)\exp{\ii k_r r}.
\end{align}
We take $k_r>0$ without loss of generality, and further assume $k_rr\gg1$ so that curvature terms can also be 
ignored.

In order to emphasize the radially local approximation, we
introduce the local Cartesian co-ordinates $(x,y,z)$ corresponding to
the the local $(r,\phi,z)$ directions, respectively. Then writing
$\delta P /\rho \equiv W$ and $c_s^2\delta\rho/\rho\equiv Q$, the
linearized system of equations are 
\begin{align}
  \ii\sigma \frac{Q}{c_s^2}  &=  \ii k_x \delta v_x + \frac{d\dd
    v_z}{dz} + \frac{\p\ln{\rho}}{\p z}\delta v_z + \hat{g}_c
  \frac{\p\ln{\rho}}{\p r} \delta v_x,\label{lin_mass}\\
  \ii\sigma \delta v_x  &= \ii k_x W - 2\Omega\delta v_y -
  \hat{g}_c\frac{1}{\rho}\frac{\p P}{\p r}\frac{Q}{c_s^2}\label{lin_vx}\\
   \ii \sigma\delta v_y &= r\frac{\p \Omega}{\p z}\delta v_z +
  \frac{\kappa^2}{2\Omega}\delta v_x, \label{lin_vy}\\
   \ii\sigma\delta v_z &= \frac{dW}{dz} +
  \frac{\p\ln{\rho}}{\p z}\left(W-Q\right), \label{lin_vz}\\
  \ii \sigma W &= c_s^2\frac{\p\ln{\rho}}{\p z}\delta v_z +
  c_s^2\frac{\gamma}{\Gamma} \left(\ii k_x\delta v_x + \frac{d\delta
      v_z}{dz}\right) \notag\\
  &+ \frac{1}{t_c}\left(W - \frac{Q}{\Gamma}\right) +
  \hat{g}_c\frac{1}{\rho}\frac{\p P}{\p r}\delta v_x,\label{lin_energy}
  % -&\ii\sigma W + \frac{1}{t_c}\left(W-\frac{Q}{\Gamma}\right) = -\ii\sigma\frac{\gamma}{\Gamma} Q +
  % c_s^2\frac{d\ln{\rho}}{dz}\left(\frac{\gamma}{\Gamma}-1\right)\delta v_z.\label{lin_energy}
\end{align}
where we have made the notation change $k_r\to k_x$. 
It is understood that the quantities $\rho,\,c_s^2,\,\Omega,\kappa,\, 
\p_z\Omega,\, \p_r\ln{\rho} $ and $ \p_r P$ are evaluated at $r=r_0$
using their expressions in the global disk given in \S\ref{setup}. Thus, they are functions of $z$ only. 
Eq. \ref{lin_mass}---\ref{lin_energy} will form the basis of our numerical
study, but for analytical discussion
we will make further simplifications.

>>>>

\section{Linear problem}\label{linear}

We consider axisymmetric perturbations to the above equilibria in the
form 
\begin{align}
  \rho \to \rho(r, z) + \delta\rho(r,z)\exp{\left( - \ii\zeta
      t\right)},    
\end{align}
and similarly for other fluid variables, where $\zeta = \omega + \ii
\sigma$ is a (generally) complex frequency, with $\omega$ and $\sigma$ being
the real frequency and growth rate, respectively.

% The linearized fluid equations in cylindrical co-ordinates are  
% \begin{align}
%   \ii\zeta\delta\rho &= \frac{\rho}{r}\frac{\p}{\p r}\left(r\delta v_r\right) + \frac{\p}{\p
%     z}\left(\rho\delta v_z\right)  + \hat{g}_c\frac{\p\rho}{\p r}\delta v_r,\label{full_mass1}\\
%   \ii\zeta\delta v_r &= \frac{1}{\rho}\frac{\p\delta P}{\p r} - 2\Omega
%   \delta v_\phi - \hat{g}_c\frac{\delta\rho}{\rho^2}\frac{\p P}{\p r},\\ 
%   \ii\zeta\delta v_\phi  &= \frac{\kappa^2}{2\Omega}\delta v_r +
%   r\frac{\p\Omega}{\p z}\delta v_z,\\
%   \ii \zeta\delta v_z &= \frac{1}{\rho}\frac{\p\delta P}{\p z} -
%   \frac{\delta \rho}{\rho^2}\frac{\p P}{\p z},\\
%   \ii\zeta\delta P &=\frac{\p P}{\p z}\delta v_z + \gamma P
%   \left[\frac{1}{r}\frac{\p}{\p r}\left(r\delta v_r\right) + \frac{\p
%       \delta v_z}{\p z}\right]\notag \\
%   &+ \frac{1}{t_c}\left(\delta P -  
%     \frac{P}{\rho}\delta\rho\right) + \hat{g}_c \frac{\p P}{\p r}\delta v_r,\label{full_energy1}
% \end{align}
% where the coefficients $\hat{g}_c=1$ are inserted to label terms
% associated with the background radial disk structure.

@@

We consider perturbations with small radial length-scales 
so that we can ignore the radial variations in the background
structure and radial gradients. \citetalias{barker15} have validated this
approximation by comparison with the linear VSI in global cylindrical
disk geometry. We thus evaluate all coefficients in
Eq. \ref{full_mass1}---\ref{full_energy1} at the fiducial radius
$r=r_0$. However, the full vertical dependence of the coefficients are
retained. Then we can assume perturbations have spatial dependence in
the form 
\begin{align}
  \delta \rho(r,z) \to \delta\rho(z)\exp{\ii k_r r}.
\end{align}
We take $k_r>0$ without loss of generality, and further assume $k_rr\gg1$ so that curvature terms can also be 
ignored.

In order to emphasize the radially local approximation, we
introduce the local Cartesian co-ordinates $(x,y,z)$ corresponding to
the the local $(r,\phi,z)$ directions, respectively. Then writing
$\delta P /\rho \equiv W$ and $c_s^2\delta\rho/\rho\equiv Q$, the
linearized system of equations are 
\begin{align}
  \ii\zeta \frac{Q}{c_s^2}  &=  \ii k_x \delta v_x + \frac{d\dd
    v_z}{dz} + \frac{\p\ln{\rho}}{\p z}\delta v_z + \hat{g}_c
  \frac{\p\ln{\rho}}{\p r} \delta v_x,\label{lin_mass}\\
  \ii\zeta \delta v_x  &= \ii k_x W - 2\Omega\delta v_y -
  \hat{g}_c\frac{1}{\rho}\frac{\p P}{\p r}\frac{Q}{c_s^2}\label{lin_vx}\\
   \ii \zeta\delta v_y &= r\frac{\p \Omega}{\p z}\delta v_z +
  \frac{\kappa^2}{2\Omega}\delta v_x, \label{lin_vy}\\
   \ii\zeta\delta v_z &= \frac{dW}{dz} +
  \frac{\p\ln{\rho}}{\p z}\left(W-Q\right), \label{lin_vz}\\
  \ii \zeta W &= c_s^2\frac{\p\ln{\rho}}{\p z}\delta v_z +
  c_s^2\frac{\gamma}{\Gamma} \left(\ii k_x\delta v_x + \frac{d\delta
      v_z}{dz}\right) \notag\\
  &+ \frac{1}{t_c}\left(W - \frac{Q}{\Gamma}\right) +
  \hat{g}_c\frac{1}{\rho}\frac{\p P}{\p r}\delta v_x,\label{lin_energy}
  % -&\ii\zeta W + \frac{1}{t_c}\left(W-\frac{Q}{\Gamma}\right) = -\ii\zeta\frac{\gamma}{\Gamma} Q +
  % c_s^2\frac{d\ln{\rho}}{dz}\left(\frac{\gamma}{\Gamma}-1\right)\delta v_z.\label{lin_energy}
\end{align}
where we have made the notation change $k_r\to k_x$. 
It is understood that the quantities $\rho,\,c_s^2,\,\Omega,\kappa,\, 
\p_z\Omega,\, \p_r\ln{\rho} $ and $ \p_r P$ are evaluated at $r=r_0$
using their expressions in the global disk given in \S\ref{setup}. Thus, they are functions of $z$ only. 
Eq. \ref{lin_mass}---\ref{lin_energy} will form the basis of our numerical
study, but for analytical discussion
we will make further simplifications.
